---
title: "Joelle_Fitzgerald_GroupCompAnalysis"
author: "Joelle Fitzgerald"
date: "10/18/2022"
output: html_document
---

## HIDS-503_2022 Midterm Project: Alzheimer’s disease study - utilizing gene expression data for discovery of new molecular biomarkers and potential drug targets 

  Each team to conduct differential gene expression analysis (in R) to find out what are the significant differences of gene expression between groups of subjects at various stages of AD in comparison with normal control.  

## Midterm Project Part 1: Gene Expression Analysis, Welch's T-Test

### Given: 
 Alzheimer patient dataset of clinical and gene expression data 
 
  * Clinical data on 30 patients. 
  * Gene expression data on 30 patients. Processed normalized gene expression data in log2 scale.
  
## Step 0 
* Open clinical data file in excel (if possible) to get familiar with the data. Run the following checks:
    + Rows are patients. 
    + Columns are clinical attributes
    + There is one row per patient
    + Check/Notice the Number of rows and number of columns
    + All data must be de-identified

* Ensure gene expression data is cleaned (no missing or inaccurate data)

## Step 1 - Read in data into R. Check if data was imported correctly
* Read clinical data into R
    + General Checks 
    + Rows are patients
    + Columns are clinical attributes
    + There is one row per patient
    + How many patients are there (dimensions of the data frame)
    + Find the column which is your outcome of interest
    + What are the groups to compare ?
* Read in cleaned processed molecular data into R
    + Check dimensions of the data frame
    + Rows are molecular features/variables (genes in our example)
    + Columns are patients
* Repeat for each molecular data type

### Step 1a - Read in clinical data

```{r}

clinData <- read.csv(file = "code/input/Blalock_clin_final.csv")
knitr::kable(head(clinData[,1:4],n = 5))

```

### Step 1b: Read in processed normalized gene expression data in log2 scale, includes gene annotation 

```{r}

geneExp <- read.csv(file = "code/input/GSE62232_Blalock_geneexp_final.tsv",
                       sep="\t", 
                       header = T, 
                       stringsAsFactors = F, row.names = 1)
knitr::kable(geneExp[1:5, 1:4], caption = "Processed gene expression data in log2 scale with gene annotation")

```

### Comparison and Baseline Groups

* Question: What is the outcome variable column from clinical data file? It is the column in the clinical file that contains the baseline and comparison values
* Answer: "Disease_Status" column has values "Control" (baseline group) and "Incipient" (comparison group)

```{r}
#Returns the value and frequency of 'Control', 'Incipient', 'Moderate', and 'Severe' in ClinData data set

knitr::kable(table(clinData$DISEASE_STATUS)) # 8 'Control', 7 'Incipient'

```

## Step 2 - Clean/Filter data
### Filter clinical data 

Clinical data has one row per patient, but only compare control and incipient samples, so we need to subset ClinData.

```{r}

subsetClinData <- clinData[which(clinData$DISEASE_STATUS=='Incipient' | clinData$DISEASE_STATUS=='Control'), ]

head(subsetClinData)
  
```

### Filter gene expression data

Check rows and columns:
* Rows are features (Disease_Status)
* Columns are patients (Biospecimen_ID)

```{r}
#Finding matching samples for baseline and comparison groups across the two data frames and assigning them to new variables 

# 30 matching samples between geneExp and clinData matching on Biospecimen_ID
matchingSamples <- which(colnames(geneExp) %in% clinData$BIOSPECIMEN_ID) 

#30 matching samples subset in geneExp
subsetGeneExp <- geneExp[, matchingSamples] 

```

## Step 3 - Identify the groups to be compared

### Identify the groups to be compared (Baseline and Comparison Grps)

```{r}
# Labels (row numbers) that can identify the baseline group patients
baselineLabels <-  which(clinData$DISEASE_STATUS == "Control") 
# 8 samples generated 
head(baselineLabels)

# Use the labels (row numbers) to subset baseline patients in clinical data file 
clinBase <- clinData[baselineLabels, ]

# Labels (row numbers) that can identify the comp group patients
compLabels <- which(clinData$DISEASE_STATUS == "Incipient") 
# 7 samples
head(compLabels)

# Use the labels (row numbers) to subset comp patients in clinical data file 
clinComp <- clinData[compLabels, ]

#### Use the clinBase and clinComp objects to subset gene expression data
geneExpBase <- subsetGeneExp[, clinBase$BIOSPECIMEN_ID] 
head(geneExpBase)
# 43135 feature (rows), 8 samples columns

geneExpComp <- subsetGeneExp[, clinComp$BIOSPECIMEN_ID] 
head(geneExpComp)
# 43135 feature (rows), 7 samples columns

```

## Step 4 - Sanity check

* See if filtering of clinical data in R matches filtering of clinical data in excel
* See if sample ids in clinical data match sample ids in gene exp data (if they don't match it means your step 1 and/or 2 is wrong)
* Verify you see correct number of samples in baseline and comp groups
* Export the column names from gene expression data to see if it contains only gene feature names and no other garbage

```{r}
#See if sample ids in clinical data match sample ids in gene exp data
clinBase$BIOSPECIMEN_ID == colnames(geneExpBase)
clinComp$BIOSPECIMEN_ID == colnames(geneExpComp)

### Checking if the samples IDs baseline and comp groups are the same ---- you can put these in an excel to see if IDs match
write.table(clinBase$BIOSPECIMEN_ID , file = "code/output/Joelle_Fitzgerald_ClinBaseIDs.tsv", sep="\t", quote = F  )
write.table(clinComp$BIOSPECIMEN_ID , file = "code/output/Joelle_Fitzgerald_ClinCompIDs.tsv", sep="\t", quote = F  )
write.table(colnames(geneExpBase) , 
            file = "code/output/Joelle_Fitzgerald_GeneExpBaseIDs.tsv", 
            sep="\t", quote = F  )
write.table(colnames(geneExpComp) , 
            file = "code/output/Joelle_Fitzgerald_GeneExpCompIDs.tsv", 
            sep="\t", quote = F  )

#Export the features from gene expression data  
#Open this file and check that it contains only gene names and no other garbage
write.table(rownames(subsetGeneExp),file = "code/output/genExpFeatureIDsCheck.tsv", sep="\t", quote = F  )

```

## Step 5 - Prep data for T-test

* Molecular data must have features (genes) as rows, and samples as columns.
* Objects must be data frame
* Numeric data only

4. CSV file results of T-test obtained from above step (obtained by sorting on p-value)
5. A text document that contains only the top 20 genes 

```{r}
### Checking to make sure data is a numeric data frame
knitr::kable(head(geneExpBase[1:5,1:4]))
knitr::kable(head(geneExpComp[1:5,1:4]))
```

#### Function for T-test

```{r ttest}
source("code/fnTTest.R")

#### Call T-test function
results1 = fnTTest(baseGroup = geneExpBase, 
                   compGroup = geneExpComp, 
                   toolN = "Joelle_Fitzgerald_TTest_", 
                   baseGroupName = "Control_",
                   compGroupName =  "Incipient_",
                   folderName = "output")
```

*The output of the T-Test analysis is a CSV File , and will be located in the folder Name that you provided in the previous step. Locate this file.*

# Next Step - Sub-set top differentially expressed genes

Cleared workspace & read in the T-Test results file.

```{r}
#Read in the T-Test results file
ttestResults <- read.csv(file = "output/Joelle_Fitzgerald_TTest__Incipient__(Comp).vs._Control__(Base).TTest.csv")

#check to make sure p-value column is imported as numeric 
#sort by p-value (just in case the results are not sorted by p-value)
ttestResultsSorted <- dplyr::arrange(ttestResults, Pvalue)

#find rows with p-value < 0.01
whichSig <- which(ttestResultsSorted$Pvalue <= 0.01) 

#Short list sig results
ttestResultsSig <- ttestResultsSorted[whichSig, ] #719 rows with 17 variables

### Export short listed results
write.table(x = ttestResultsSig, 
            file = "output/Joelle_Fitzgerald_Control_Incipient_Ttest_Shortlisted.tsv", 
            quote = F, sep = "\t")

##### First column is a list of features in this format : ID|Disease_Status. 
#### Use string split strsplit() function to extract gene names
funcSplit <- function(featureX) {
  f1 <- unlist(strsplit(x = featureX, split = "|", fixed = TRUE))
  f2 <- f1[2]
  return(f2)
}

# Use apply() function to run the split on every row, its faster version of a loop
Status1 <- apply(X = as.matrix(ttestResultsSig$Feature), 
                    MARGIN = 1, FUN = funcSplit)

head(Status1) # Visual of split function results for disease status & corresponding genes

#print length of short listed gene names,  results
length(Status1)

### Export list of top 20 gene names
write.table(x = head(Status1,20), 
            file = "output/Joelle_Fitzgerald_SigDiff_DiseaseStatus.tsv", 
            quote = F, sep = "\t")

 
```

Next, we will do Enrichment Analysis in Part 2. 

